
The fast Fourier transform (FFT) has revolutionized a wide range of applications
in mathematics, engineering, physics, etc., by enabling signal processing and
Fourier analysis tasks to be performed using a computational cost which scales
quasi-linearly with the number of data points~$n$. However, the FFT requires
that the input signal be sampled at equispaced points in time and that the
desired output frequencies are equispaced on the integers. These assumptions are
frequently not met in applications such as adaptive numerical PDE
solvers~\cite{alpert2002adaptive,jiang2023dual,askham2017adaptive,nochetto2009theory},
magnetic resonance
imaging~\cite{greengard2007fast,bondesson2019nonuniform,bronstein2002reconstruction},
and various signal processing
tasks~\cite{alexander2012adaptive,thakur2011synchrosqueezing}. To overcome this
setback, nonuniform FFT (NUFFT) algorithms have been
developed~\cite{dutt1993fast,greengard2004accelerating} which achieve near-FFT
speeds in one dimension, assuming that the distribution of time samples and
frequency outputs is not pathological. In higher dimensions, NUFFTs are less
competitive with standard FFTs, but the computational task at hand is also
significantly harder.

The FFT grew out of a need to perform Fourier transforms in Cartesian
coordinates. However, depending on the particular problem, the relevant
continuous Fourier analysis might be better suited to other coordinate systems.
One such commonly encountered situation is computing the Fourier transform of
radially symmetric functions in dimensions~$d \geq 2$. For example, in two
dimensions the Fourier transform of a function~$f$ is given by
\begin{equation}
  g(\omega_1, \omega_2) = \frac{1}{4\pi^2} \iint_{\R^2} f(x_1, x_2) \, 
  e^{-i(\omega_1 x_1 + \omega_2 x_2)}  \, dx_1 \, dx_2.
\end{equation}
Transforming to polar coordinates~$(\omega_1,\omega_2) \mapsto (\omega,\alpha)$
and~$(x_1,x_2) \mapsto (r,\theta)$ the above expression becomes
\begin{equation}
  \label{eq:ftpolar}
  \begin{aligned}
    g(\omega, \alpha) &= \frac{1}{4\pi^2} \int_0^{2\pi} \int_0^\infty
    f(r,\theta) \, 
    e^{-i \omega r (\cos\alpha \cos\theta + \sin\alpha \sin\theta) } 
    \, r \, dr \, d\theta \\
  &= \frac{1}{4\pi^2} \int_0^{2\pi} \int_0^\infty f(r,\theta) \, e^{-i \omega r \cos(\alpha-\theta) } \, r \, dr \, d\theta.
  \end{aligned}
\end{equation}
Furthermore, if~$f$ is radially symmetric, i.e.~$f(r, \theta) = f(r)$, then the
above transform can be written as
\begin{equation}
  \label{eq:HT}
  \begin{aligned}
  g(\omega,\alpha) &= \frac{1}{4\pi^2} \int_0^\infty f(r) \, r \int_0^{2\pi} 
  e^{-i \omega r \cos(\alpha - \theta) }  \, d\theta \, dr \\
  &= \frac{1}{2\pi} \int_0^\infty f(r) \, J_0(\omega r) \, r \, dr,
  \end{aligned}
\end{equation}
where we have used the integral representation of the zeroth-order Bessel
function~\cite{olver2010nist}
\begin{equation}
  J_0(x) 
  = \frac{1}{\pi} \int_0^\pi \cos \left( x \cos \theta \right) \, d\theta.
\end{equation}
The final integral involving~$J_0$ in equation~\eqref{eq:HT} is known as a
\emph{Hankel Transform} of order 0 --- usually referred to simply as a Hankel
Transform. 

In higher ambient dimensions, the Fourier transform of radially symmetric
functions reduces to a Hankel transform of higher order, which we treat toward
the end of this manuscript. Similarly, if the function~$f$ in~\eqref{eq:ftpolar}
has a particular periodic dependence in~$\theta$, e.g.~$f(r,\theta) =
f(r)e^{ik\theta}$, then we have
\begin{equation} \label{eq:FB-integral}
  \begin{aligned}
  g(\omega,\alpha) &= \frac{1}{4\pi^2} \int_0^\infty f(r) \, r \int_0^{2\pi} 
  e^{-i \omega r \cos(\alpha - \theta) } \, e^{ik\theta}  \, d\theta \, dr \\
  &= \frac{i^k}{2\pi} \int_0^\infty f(r) \, r \, J_k(\omega r)  \, dr,
  \end{aligned}
\end{equation}
where, again, we have invoked an integral representation
for~$J_k$~\cite{olver2010nist}. 

In order to numerically compute~$g$ in~\eqref{eq:HT} or~\eqref{eq:FB-integral}
at a collection of~$m$ ``frequencies''~$\omega_j$, the Hankel transform must be
discretized using an appropriate quadrature rule which depends on the particular
class of~$f$ for which the integral is desired. In general this results in the
need for computing
\begin{equation} \label{eq:DHT}
  \begin{aligned}
  g(\omega_j) \approx 
  g_j &:= \sum_{k=1}^n w_k \, f(r_k) \, r_k \, J_\nu(\omega_j r_k) \\
  &\ = \sum_{k=1}^n c_k \, J_\nu(\omega_j r_k)
   \qquad \text{for } j = 1, \ldots, m.
  \end{aligned}
\end{equation}
The above sum will be referred to as the Discrete Hankel Transform (DHT) of
order $\nu$. 

In our motivating example --- computing the continuous Fourier transform --- the
DHT arises from the discretization of the radially symmetric Fourier integral.
The DHT also appears in a wide range of applications including
imaging~\cite{higgins1988hankel, zhao2013fourier, marshall2023fast},
statistics~\cite{lord1954a, genton2002nonparametric}, and separation of
variables methods in partial differential
equations~\cite{bisseling1985fast,ali1999generalized, zhou2022spectral}. In many
such applications, a fully nonuniform DHT is desired, as the relevant
frequencies $\omega_j$ may not be equispaced, and the most efficient quadrature
rule for discretizing (\ref{eq:HT}) may have nodes $r_k$ which are also not
equispaced. It is enough to consider the above sum merely with the
coefficients~$c_k$. The algorithm of this work allows for arbitrary selection of
the frequencies~$\omega_j$ and nodes~$r_k$, in contrast to other algorithms
which require some structure to their location (e.g. equispaced). There are a
few types of commonly encountered DHTs, all of which our algorithm can address.
Schl\"omilch expansions~\cite{linton2006schlomilch,townsend2015fast}
take~$r_k \in [0,1]$ and the frequencies are chosen as~$\omega_j =
j\pi$. Fourier-Bessel expansions, often used in separation of variables
calculations for PDEs, take the frequencies~$\omega_j$ to be the ordered roots
of~$J_0$. In the most restrictive case, as discussed in~\cite{johnson1987improved}, the
frequencies~$\omega_j$ are set to be the ordered roots of~$J_\nu$, and the
nodes~$r_k$ a re set to be scaled roots of~$J_\nu$,
i.e.~$r_k = \mathrm{j}_{\nu,k}/\mathrm{j}_{\nu,k+1}$ where~$\mathrm{j}_{\nu,k}$
is the~$k$th largest root of~$J_\nu$ (which doesn't explicitly appear in the sum
above). Analogues of all these special cases exist when~$J_0$ is replaced with a
higher order bessel function~$J_m$ as well, but we do not go into the details as
our algorithm handles them all similarly.

\subsection*{Existing methods}
\label{sec:existing}

A number of methods exist in the literature to evaluate (\ref{eq:HT}) and
(\ref{eq:DHT}). These include series expansion methods
\cite{lord1954b,brunol1977fourier,cavanagh1979numerical}, convolutional
approaches \cite{siegman1977quasi, johansen1979fast, mook1983algorithm}, and
projection-slice or Abel transform-based methods \cite{oppenheim1980computation,
hansen1985fast, kapur1995algorithm}. See~\cite{cree1993algorithms} for a review
of many of these early computational approaches. Unfortunately, these existing
methods are either not applicable to the discrete case, require a particular
choice of $\omega_j$ or $r_k$ due to the constraints of interpolation or
quadrature subroutines, or suffer from low accuracy as a result of intermediate
approximations. Therefore, extending these schemes to compute the fully
nonuniform DHT with controllable accuracy is not straightforward.

A notable contribution is~\cite{liu1999nonuniform}, which describes a fully
nonuniform fast Hankel transform, although it has accuracy and speed
limitations. This work takes the popular convolutional approach, using a change
of variables to reformulate the Hankel transform as a convolution with a known
kernel. However, its accuracy is limited by the need for a quadrature on the
nonuniform points $r_k$ in order to compute the continuous Fourier transform.
The authors use an irregular trapezoidal rule for this purpose, which is not
high-order accurate. This method also requires the computation of the inverse
NUFFT using conjugate gradients. For even moderately clustered points or
frequencies, this inverse problem is extremely ill-conditioned, and thus the
number of required iterations can be prohibitive. This method is therefore
suitable for ``quasi-equispaced'' points and frequencies, but is not tractable
for all distributions.

More recently, butterfly algorithms \cite{oneil2010algorithm, li2015butterfly,
pang2020interpolative} were introduced as a broadly applicable methodology for
rapidly computing oscillatory transforms including the nonuniform DHT. However,
these algorithms require a precomputation or factorization stage for each new
set of $\omega_j$ and $r_k$. Such precomputations can, unfortunately, be a
bottleneck for applications in which these evaluation points change with each
iteration or application of the transform. In order to provide a
precomputation-free fast DHT,~\cite{townsend2015fast} employs a combination of
asymptotic expansions and Bessel function identities evaluated using the
equispaced FFT. The resulting scheme is applicable to equispaced or perturbed
``quasi-equispaced'' grids, for example $j_{0,k} / j_{0,n+1}$ where $j_{\nu,k}$
is the $k^{th}$ zero of $J_\nu$.

\subsection*{Novelty of this work}
\label{sec:novelty}

% At a high level, our algorithm can be viewed as a generalization of the one
% described in~\cite{townsend2015fast}. In~\cite{townsend2015fast}, asymptotic
% expansions were used to replace~$J_0$ for various arguments. These asymptotic
% expansions involved trigonometic functions, resulting in a fast algorithm for
% computing the DHT using fast cosine transforms (FCTs) and fast sine transforms
% (FSTs). In order to invoke these fast algorithms, various assumptions
% on~$\omega_j$ and~$r_k$ had to be made.

We describe here a precomputation-free nonuniform fast Hankel transform (NUFHT)
which generalizes~\cite{townsend2015fast} to the fully nonuniform setting in a
number of ways. First, we employ an adaptive partitioning scheme which, for any
choice of $\omega_j$ and $r_k$, subdivides the matrix with
entries~$J_\nu(\omega_j r_k)$ into blocks for which matrix-vector products can
be evaluated efficiently. Second, we use the nonuniform fast Fourier transform
(NUFFT) to evaluate asymptotic expansions for nonuniform $r_k$ and $\omega_j$.
Finally, we utilize the low-rank expansion of $J_\nu$ given
in~\cite{wimp1962polynomial} in the local regime where asymptotic expansions are
not applicable. We derive error bounds for this low-rank expansion, allowing us
to choose all approximation parameters automatically by analysis which
guarantees that the resulting error is bounded by the user-specified tolerance
$\epsilon$.

\subsection*{Outline of the paper}

The paper is organized as follows. In Section~\ref{sec:overview} we give a high
level view of our algorithm, omitting technical details. Then in
Section~\ref{sec:approx} we study the local and asymptotic expansions of Bessel
functions which serve as the key building blocks of the algorithm. Afterward, in
Section~\ref{sec:methods}, we provide a detailed description of the algorithm
and its associated complexity. Various numerical examples are provided in
Section~\ref{sec:results}, and we conclude with some additional discussion in
Section~\ref{sec:discussion}.




%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../main"
%%% End:
